Popualtion Dynamics of an Endangered Frog Recovering from Disease-Driven Decline

### 2025-05-07 code/R/Survival_by_sex.Rmd
organize R code files

@@

# Create the plot
difference_plot <- ggplot(time_summaries_df, aes(x = primary_period, y = mean)) +
  geom_point(color = "blue", size = 2) +  # Increased point size for better visibility
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.2, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Difference in Survival Probabilities (M - F) Between Primary Periods\nsurvival_fill_value = 'M'",
    x = "Primary Period",
    y = "Difference in Survival Probability"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Tilt x-axis labels for readability
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")  # Center and style the title
  ) +
  # Correct the x-axis limits to match 'primary_period' labels
  scale_x_discrete(limits = levels(time_summaries_df$primary_period))

@@

# Save the plot to a file
ggsave("/Users/johnimperato/Desktop/Mossy-Pond_companion/difference_in_survival_probabilities_M_fill.png", 
       plot = difference_plot, width = 10, height = 6, dpi = 300, bg = "white")
```

>>>>

write_csv(time_summaries_df, "/Users/johnimperato/Downloads/MF_survival.csv")

@@

# Create the plot
difference_plot <- ggplot(time_summaries_df, aes(x = primary_period, y = mean)) +
  geom_point(color = "blue", size = 2) +  # Increased point size for better visibility
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.2, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Primary Period",
    y = "Difference in Survival Probability (M-F)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Tilt x-axis labels for readability
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")  # Center and style the title
  ) +
  # Correct the x-axis limits to match 'primary_period' labels
  scale_x_discrete(limits = levels(time_summaries_df$primary_period))

@@

# Save the plot to a file
ggsave("/Users/johnimperato/Desktop/Mossy-Pond_companion/difference_in_survival_probabilities_ratio_fill.png", 
       plot = difference_plot, width = 10, height = 6, dpi = 300, bg = "white")
```

library(ggplot2)
library(dplyr)

# 1. Prepare transition labels
time_summaries_df <- time_summaries_df %>%
  arrange(time_index) %>%
  mutate(
    transition = factor(
      paste0(time_index, "-", time_index + 1),
      levels = paste0(time_index, "-", time_index + 1)
    )
  )

# 2. Compute y-axis lower bound so that zero sits above the bottom 
y_lower <- min(time_summaries_df$lower95, 0) - 0.1
y_upper <- max(time_summaries_df$upper95) + 0.01

# 3. Plot
difference_plot <- ggplot(time_summaries_df, aes(x = transition, y = mean)) +
  geom_point(size = 2, color = "black") +
  geom_errorbar(aes(ymin = lower95, ymax = upper95),
                width = 0.2, color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#B2182B") +
  scale_y_continuous(
    limits = c(y_lower, y_upper),
    expand = c(0, 0)
  ) +
  labs(
    x = "Primary Period Transitions",
    y = "Difference in Survival Probability (M–F)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title    = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# 4. Display
print(difference_plot)

### 2025-05-07 code/R/WinterSeverity_analysis_MossyPond.Rmd
organize R code files

avg_swe <- avg_swe %>%
  rename(Winter_Severity = Average.Pixel.Value) %>%
  mutate(Winter = c("2010-2011", "2011-2012", "2012-2013", "2013-2014", "2014-2015", "2015-2016", "2016-2017", "2017-2018"))

@@

## Population Vital Rates/Numbers and Growth Rate

@@

model_summary <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_custom_mod_sumary.csv')

@@

@@

# extract overwinter survival values from model summary
phi <- B <- model_summary[c(33738:33752), ]

@@

print(N)
print(B)

@@

@@

ggsave("WinterSeverityB1.png", plot = WinterSeverity_B1, bg = "white")
ggsave("CorrelationB1.png", plot = correlation_plot_B1, bg = "white")

@@

correlation_plot_B3 <- ggplot(data = comparison_df_B3, aes(x = swe_3yrAvg, y = Recruited_Individuals)) +
  geom_point() +
  geom_smooth(method = "lm" ,se = FALSE) +
  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_B3, 2)), size = 5, color = "red") +
  labs(
    title = "Correlation between 3-yr avg Winter Severity and Recruitment",
    x = "Winter Severity (% of Historical Average 1 April SWE)",
    y = "Yearly Mean Abundance"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

>>>>

avg_swe <- avg_swe %>%
  mutate(Winter = c("2010-2011", "2011-2012", "2012-2013", "2013-2014", "2014-2015", "2015-2016", "2016-2017", "2017-2018"))

@@

## Population Vital Rates/Numbers

@@

model_summary <- read.csv('/Users/johnimperato/Desktop/Mossy-Pond_companion/mp_custom_mod_summary.csv')

@@

View(N)

@@

# extract overwinter survival values from model summary
phi <- model_summary[c(33738:33752), ]
View(phi)

@@

@@

abundance_WinterSeverity_2yr
correlation_plot_2_N

@@

ggsave("WinterSeverityB1.png", plot = WinterSeverity_B1, bg = "white")
ggsave("CorrelationB1_narrow.png", plot = correlation_plot_B1, bg = "white")

@@

correlation_plot_B3 <- ggplot(data = comparison_df_B3, aes(x = swe_3yrAvg, y = Recruited_Individuals)) +
  geom_point() +
  geom_smooth(method = "lm" ,se = FALSE) +
  annotate("text", x = 75, y = 185, label = paste0("Correlation: ", round(correlation_B3, 2)), size = 5, color = "red") +
  labs(
    title = "Correlation between 3-yr avg Winter Severity and Recruitment",
    x = "Winter Severity (% of Historical Average 1 April SWE)",
    y = "Yearly Recruitment"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

### 2025-05-07 doc/Manuscript.qmd
organize R code files

---
title: Dynamics of an Endangered Frog Population Recovering from Bd-Driven Decline
format:
  pdf:
    documentclass: article
    linestretch: 2
    number-sections: False
    geometry: margin=1in
    include-in-header:
      text: |
        \usepackage{lineno}
        \linenumbers
bibliography: 
  - references.bib
  - manual.bib
execute:
  echo: false
---

@@

Mountain yellow-legged frogs (*Rana sierrae* and *R. muscosa*) have disappeared from \>90% of their historic range due to *Batrachochytrium dendrobatidis* (Bd) infection, yet a handful of populations now persist and recover in the pathogen's presence. We studied one such recovering population at Mossy Pond, Tahoe National Forest, for five summers (2014–2018) using robust-design capture–mark–recapture methods and a Bayesian Jolly–Seber model. From 3,631 captures of 657 uniquely marked adults, estimated abundance varied from 117 to 245 individuals, while 629 new adults recruited over the study. Monthly survival was strongly seasonal—mean 0.97 in winter versus 0.80 in summer—with male survival exceeding female survival by 2–7% in each interval. Survey-specific detection probability averaged 0.45. Bd was present in 67% of skin-swab samples and infection loads remained two–three orders of magnitude below lethal thresholds. Recruitment and abundance declined sharply after severe winters, yet overwinter adult survival showed no snow effect. These findings confirm functional Bd resistance, reveal contrasting seasonal pressures that shape population dynamics, and identify this robust northern Sierra Nevada population as a prime donor for translocation-based recovery efforts.

@@

Amphibians continue to experience global declines, with infectious diseases recognized as primary drivers of biodiversity loss [@luedtke2023a]. Chytridiomycosis, caused by the fungal pathogen *Batrachochytrium dendrobatidis* (Bd), has driven declines or extinctions of approximately 200 amphibian species globally [@skerratt2007]. Amphibian responses to Bd infection vary dramatically; some populations show minimal impact, others rapidly go extinct, and a rare few exhibit persistent recovery following severe declines [@lips2016]. The mechanisms supporting persistence or recovery remain poorly understood but likely involve interactions among ecological, evolutionary, immunological, and demographic factors [@scheele2017].

The mountain yellow-legged (MYL) frog species complex (*Rana sierrae* and *R. muscosa*), endemic to California's Sierra Nevada and Transverse mountain ranges, provides a particularly well-documented case of amphibian decline and recovery from Bd infection. Starting in the 1970s, Bd spread through the Sierra Nevada, causing widespread and rapid extirpations of MYL frog populations [@vredenburg2010; @vredenburg2019]. By the early 2000s, over 90% of their historical range had been lost, due to combined impacts from Bd and habitat degradation caused by introduced trout [@knapp2000]. Despite these losses, a few populations survived initial Bd outbreaks, eventually exhibiting increased abundance and apparent disease resistance [@knapp2016]. These dynamics exemplify patterns of evolutionary rescue [@carlson2014], in which populations rapidly acquire adaptive genetic variation that restores demographic viability under persistent threats [@knapp2024].

While broad-scale studies have described these declines and recoveries, detailed demographic analyses at the individual population level remain scarce, primarily due to logistical challenges. Most fine-scale demographic research has focused on southern populations, particularly within Yosemite National Park [@fellers2013; @joseph2018; @knapp2024] . In contrast, the population dynamics of frogs in the northern Sierra Nevada—where habitats are more heterogeneous and populations are smaller and lower in elevation—remain understudied (MYLF ITT 2018) [@brown2019].

To address this gap, we conducted a five-year capture-mark-recapture (CMR) study of a naturally recovering, Bd-persistent *R. sierrae* population in Tahoe National Forest. Our objectives were to estimate key demographic parameters—abundance, survival, recruitment, and detection probability—to provide insights into the ecological processes and factors that contribute to population viability under endemic Bd infection. We also examined the role of climate variability in shaping these demographic parameters, as climate change is increasingly recognized as a primary driver of amphibian declines worldwide, comparable in impact to disease and habitat loss [@luedtke2023a]. Understanding how climate influences abundance, survival, and recruitment is particularly important for high-elevation amphibian populations living with endemic Bd infection.

@@

The Mossy Pond study area is a one square mile section of the Tahoe National Forest, north of Highway 80 in Nevada County, California. It is characterized by lakes, ponds, and streams set on granite benches, ranging from 6,400 feet to 7,100 feet in elevation. The study area includes the namesake Mossy Pond (6 hectare surface area, maximum depth 2.5 meters), its seasonally flowing outlet stream, and 12 ephemeral ponds. The absence of deep-water habitat is unusual for areas supporting MYL frog populations, given their reliance on water bodies deeper than 4 meters for overwinter survival [@bradford1983; @knapp2003].

*R. sierrae* is a federally endangered, mid-sized ranid frog that occupies montane waterbodies in California's Sierra Nevada [@u.s.fishandwildlifeservice2014]. The US Geological Survey first detected *R. sierrae* in the Mossy Pond area in 1998, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.

@@

To assess the Mossy Pond *R. sierrae* population's Bd status, we collected epithelial swabs from 24 adult frogs over three years (2008, 2010, and 2021). Partner scientists quantified the amount of Bd DNA on each swab using real-time quantitative polymerase chain reaction (qPCR) analysis [@boyle2007]. We report infection loads as the number of ITS1 (Internal Transcribed Spacer 1) gene copies detected [@longo2013; @joseph2018]. Quantifying Bd loads rather than simply testing for Bd presence allowed us to compare our study population with other known recovering populations. Partner scientists provided *R. sierrae* Bd load data from three known naturally recovering populations in Yosemite National Park [@knapp2024].

@@

We studied *R. sierrae* with capture-mark-recapture (CMR) methods, which are widely used for estimating wildlife population parameters such as abundance, survival, and recruitment [@williams2001]. We conducted CMR sampling according to Pollock's robust design [@pollock1982], which involves sampling at two temporal scales: primary periods (between which the population is open to demographic change) and secondary periods, during which the population is assumed closed. Each summer from 2014 to 2018, we visited the study site 3 or 4 times (primary periods) for three consecutive days of surveys (secondary periods), for a total of 47 individual surveys.

>>>>

---
title: Popualtion Dynamics of an Endangered Frog Recovering from Disease-Driven Decline
format:
  pdf:
    documentclass: article
    linestretch: 2
    number-sections: False
    geometry: margin=1in
    include-in-header:
      text: |
        \usepackage{lineno}
        \linenumbers
bibliography: 
  - references.bib
  - manual.bib
execute:
  echo: false
---

@@

Mountain yellow-legged frogs (*Rana sierrae* and *Rana muscosa*) have disappeared from \>90% of their historic range due to the virulent fungal pathogen *Batrachochytrium dendrobatidis* (Bd), yet a handful of populations now persist and recover in the pathogen's presence. We studied one such recovering population at Mossy Pond, Tahoe National Forest, for five summers (2014–2018) using robust-design capture–mark–recapture methods and analyzed the capture histories with a Bayesian Jolly–Seber model. From 3,631 captures of 657 uniquely marked adults, estimated abundance varied from 117 to 245 individuals, and a total of 629 new adults recruited over the study. Monthly survival was higher in winter than in summer (mean and credible interval; winter: 0.97 (xx-xx), summer: 0.8 (xx-xx) and male survival exceeded female survival by 2–7% in each interval. Survey-specific detection probability averaged 0.45. Bd was present in 67% of skin-swab samples and infection loads were two–three orders of magnitude below lethal thresholds. Recruitment and abundance declined sharply after severe winters, yet overwinter adult survival showed no snow effect. These findings suggest this population is persisting despite ongoing Bd infection, perhaps as a result of some degree of infection resistance. ................. identify this robust northern Sierra Nevada population as a prime donor for reintroduction-based recovery efforts.

@@

Amphibians continue to experience global declines, with infectious diseases recognized as primary drivers of biodiversity loss [@luedtke2023a]. Chytridiomycosis, caused by the fungal pathogen *Batrachochytrium dendrobatidis* (Bd), has driven declines or extinctions of approximately 200 amphibian species globally [@skerratt2007]. Amphibian responses to Bd infection vary dramatically; some populations show minimal impact, others rapidly go extinct, and a rare few exhibit persistence or even recovery following severe declines [@lips2016]. The mechanisms supporting persistence or recovery remain poorly understood but likely involve interactions among ecological, evolutionary, immunological, and demographic factors [@scheele2017].

The mountain yellow-legged (MYL) frog species complex (*Rana sierrae* and *Rana muscosa*), endemic to California's Sierra Nevada and Transverse ranges, provides a particularly well-documented case of amphibian decline and recovery from Bd infection. Starting in the 1970s, Bd spread through the Sierra Nevada, causing widespread and rapid extirpations of MYL frog populations [@vredenburg2010; @vredenburg2019]. By the early 2000s, over 90% of historical populations had been lost, due to combined impacts from Bd and introduced predatory trout [@knapp2000]. Despite these losses, a few populations survived initial Bd outbreaks, eventually exhibiting increased abundance and apparent disease resistance [@knapp2016]. These dynamics are consistent with patterns of evolutionary rescue [@carlson2014], in which populations rapidly acquire adaptive genetic variation that restores demographic viability under persistent threats [e.g. @knapp2024].

Although broad-scale studies have described these declines and recoveries, detailed demographic analyses at the individual population level remain scarce, primarily due to logistical challenges. Most fine-scale demographic research has focused on southern populations, particularly within Yosemite National Park [@fellers2013; @joseph2018; @knapp2024] . In contrast, the population dynamics of *R. sierrae* in the northern Sierra Nevada—where habitats are more heterogeneous and populations are smaller and lower in elevation—remain understudied (MYLF ITT 2018) [@brown2019].

To address this gap, we conducted a five-year capture-mark-recapture (CMR) study of a naturally recovering, Bd-positive *R. sierrae* population in Tahoe National Forest. Our objectives were to estimate key demographic parameters—abundance, survival, recruitment, and detection probability—to provide insights into the ecological processes and factors that contribute to population viability under endemic Bd infection. We also examined the role of climate variability in shaping these demographic parameters, as climate change is increasingly recognized as a primary driver of amphibian declines worldwide, comparable in impact to disease and habitat loss [@luedtke2023a]. Understanding how climate influences abundance, survival, and recruitment is particularly important for high-elevation amphibian populations living with endemic Bd infection.

@@

The Mossy Pond study area is a 2.6 square km section of the Tahoe National Forest, north of Highway 80 in Nevada County, California. It is characterized by lakes, ponds, and streams set on granite benches, ranging from 1,950 m to 2,164 m in elevation. The study area includes the namesake Mossy Pond (6 hectare surface area, maximum depth 2.5 meters), its seasonally flowing outlet stream, and 12 ephemeral ponds. The absence of deep-water habitat is unusual for areas supporting MYL frog populations, given their general reliance on water bodies deeper than 3-4 meters for overwinter survival [@bradford1983; @knapp2003].

*R. sierrae* is a federally endangered, mid-sized ranid frog that occupies montane waterbodies in California's Sierra Nevada [@u.s.fishandwildlifeservice2014]. *R. sierrae* was first detected in the Mossy Pond area in 1998 **(citation)**, and the California Department of Fish and Wildlife (CDFW) has been monitoring the population since 2001.

@@

To assess the Mossy Pond *R. sierrae* population's Bd status, we collected skin swabs from 24 adult frogs over three years (2008, 2010, and 2021). The amount of Bd DNA on each swab was quantified using real-time quantitative polymerase chain reaction (qPCR) analysis [@boyle2007]. We report infection loads as the number of ITS1 (Internal Transcribed Spacer 1) gene copies detected [@longo2013; @joseph2018]. To compare the Bd loads of Mossy Pond frogs to those of other known recovering *R. sierrae* populations, we compared Bd loads from Mossy Pond to those from three known naturally recovering populations in Yosemite National Park [@knapp2024].

@@

We surveyed the Mossy Pond *R. sierrae* population using capture-mark-recapture (CMR) methods, which are widely used for estimating wildlife population parameters such as abundance, survival, and recruitment [@williams2001]. We conducted CMR sampling according to Pollock's robust design [@pollock1982], which involves sampling at two temporal scales: primary periods, between which the population is open to demographic change, and secondary periods, during which the population is assumed closed. Each summer from 2014 to 2018, we visited the study site 3 or 4 times (primary periods) for three consecutive days of surveys (secondary periods), for a total of 47 individual surveys.
